### 2022-08-08 main/chapter_2ljets.tex
sg checks

Our first curiosity is that the reported $S_{\mathrm{obs}}^{95} = 3.0$
for DR-High is rounded up from a result of $2.93$ that was evaluated with these
simulation methods.
This makes sense --- the limit for $n=0$ should arguably always be $3$ since
if a signal adds $\mu$ events, then the Poisson likelihood for zero data always
scales by $e^{-\mu} \approx 0.05$, and so any likelihood ratio between
background-only and signal-plus-background models is $0.05$ for $\mu = 3$.
By taking a ratio of $p$-values, $\cls$ approximates that likelihood ratio:
we see from Figure~\ref{fig:2ljets_test_stat_distributions_drhigh} that the
observed data does land in a spike at the end of the distribution that
corresponds to to pseudo-data with $n=0$; the peaks to its left correspond
to $n=1$, $n=2$ and so on, with widths due to the other parameters varied in
toy sampling.

@@

The $n=0$ case has no visible width since its likelihood is linear
(adding signal just adds a constant to the $-\log$ likelihood objectives being
minimized), but its distribution is not quite a delta function since the test
statistic depends on the noisy termination of numerical optimization
algorithms.
Resulting $\clb$ and $\clsb$ values depend on this effect ---
with $3.9$ expected background events and zero data we expect
$\clb = \mathrm{Poisson}(0\mid 3.9) = e^{-3.9} \approx 2.0\times 10^{-2}$.
And that is $15$ times larger than the evaluated $\clb = 1.3\times 10^{-3}$
because, as shown in Figure~\ref{fig:2ljets_test_stat_distributions_drhigh_n0},
the observed test statistic happened to fall quite far to the right
of that $n=0$ noise distribution.
This small $\clb$ value could appear to be an anomalously large significance
if that $p$-value were  misinterpreted as a measure of compatibility with the
background model.
Thankfully it is remedied by the $\cls$-prescribed ratio with $\clsb$,
since both $p$-values suffer from similar effects.
Statistical precision is harmed, however --- although $1$m samples should
give us around $e^{-7.9} \times 10^6 \approx 1000$ signal samples with
$n=0$, only $57$ toy samples land to the right of the observed test statistic.
And with $\approx1/\sqrt{57}$ relative statistical precision, that easily
explains the difference between limits at $3.0$ and $2.93$ despite the huge,
and likewise expensive, toy sample size.

Although I have found no \atlas\ search reporting a discovery region limit
below $3$, there are published examples with zero data but larger observed
limits~\cite{SUSY-2018-36, atlas_susy_strong_0l}.
The most extreme is $\mathrm{SR}3\ell_{[90,110]}$
of~\cite{SUSY-2018-36}, which reports a background expectation of
$1.08 \pm 0.18$ and expected limits of
$S_{\mathrm{exp}}^{95} = 3.8^{+1.3}_{-0.5}$, but zero observed data with a
whopping $S_{\mathrm{obs}}^{95} = 4.8$.
Using their \pyhf\ models published through \hepdata\ I have failed to
reproduce this effect outside of the \histfitter-\roofit ecosystem, but would
not be surprised if similar complications with the test statistic distribution
played a part.

Not only is that limit more than $3$, it is more than the expected limit
despite the fact that increasing data should increase the observed limit.
The same effect has occurred in our results for DR-OffShell --- its observation
of $21$ data is below the post-fit background is $22.1$, but its observed limit
exceeds the central expected limit.
Another public example of this effect is
$\mathrm{SR}0\textrm{-}\mathrm{ZZ^{tight}_{bveto}}$ of~\cite{SUSY-2018-02}.
\emph{Thanks to Tina~Potter discussions of this problem and for sharing
its details of its modelling.}

Internal discussions of this error led me to perform numerous validation
studies that reproduced this finding with \pyhf\ and with various alternative
constructions of the models.
Various explanations were suggested in these discussions, including problems
with asymmetric systematic variations or asymmetries in the Poisson
distribution.
Fundamentally, a sufficient explanation of this problem is our refusal to
compare models with likelihood ratios.
By instead using \emph{maximum} likelihood ratios, and modifications thereof,
we guarantee that our tests can be suboptimal because they are not protected by
the Neyman–Pearson lemma~\cite{neyman1933ix}.
In this world of model comparison through $p$-values, sub-optimality means
that the test statistic puts some data in the wrong
order~\cite{cousins2008evaluation},
and so by knowingly using sub-optimal test statistics we also know that the
results need not behave as if the data had a sensible ordering.
Limits moving in the wrong directions can simply be explained by flawed
test statistics ranking data in a bad order.

@@

It is undoubtedly arguable that the pathological test statistic distributions
observed in DR-Low indicate a bug that should be fixed.
However, the statistical software packages that generated this result
--- \roofit\ through \histfitter ---
define in silicon the standard statistical methods for \atlas\ SUSY searches.
To diverge from them would make our result incomparable.
What they do is the definition of correct.
All \pyhf\ can do is imitate.
There is no thing as a bug.

>>>>

Our first curiosity is that the reported $S_{\mathrm{obs}}^{95} = 3.0$
for DR-High is rounded up from a result of $2.93$ that was evaluated with these
simulation methods.
This makes sense --- the limit for $n=0$ should arguably always be $3$ since
if a signal adds $\mu$ events, then the Poisson likelihood for zero data always
scales by $e^{-\mu} \approx 0.05$, and so any likelihood ratio between
background-only and signal-plus-background models is $0.05$ for $\mu = 3$.
By taking a ratio of $p$-values, $\cls$ approximates that likelihood ratio:
we see from Figure~\ref{fig:2ljets_test_stat_distributions_drhigh} that the
observed data does land in a spike at the end of the distribution that
corresponds to pseudo-data with $n=0$; the peaks to its left correspond
to $n=1$, $n=2$ and so on, with widths due to the other parameters varied in
toy sampling.

@@

The $n=0$ case has no visible width since its likelihood is linear
(adding signal just adds a constant to the $-\log$ likelihood objectives being
minimized), but its distribution is not quite a delta function since the test
statistic depends on the noisy termination of numerical optimization
algorithms.
Resulting $\clb$ and $\clsb$ values depend on this effect ---
with $3.9$ expected background events and zero data we expect
$\clb = \mathrm{Poisson}(0\mid 3.9) = e^{-3.9} \approx 2.0\times 10^{-2}$.
And that is $15$ times larger than the evaluated $\clb = 1.3\times 10^{-3}$
because, as shown in Figure~\ref{fig:2ljets_test_stat_distributions_drhigh_n0},
the observed test statistic happened to fall quite far to the right
of that $n=0$ noise distribution.
This small $\clb$ value could appear to be an anomalously large significance
if that $p$-value were  misinterpreted as a measure of compatibility with the
background model.
Thankfully, it is remedied by the $\cls$-prescribed ratio with $\clsb$,
since both $p$-values suffer from similar effects.
Statistical precision is harmed, however --- although $1$m samples should
give us around $e^{-7.9} \times 10^6 \approx 1000$ signal samples with
$n=0$, only $57$ toy samples land to the right of the observed test statistic.
And with $\approx1/\sqrt{57}$ relative statistical precision, that easily
explains the difference between limits at $3.0$ and $2.93$ despite the huge,
and likewise expensive, toy sample size.

Although I have found no \atlas\ search reporting a discovery region limit
below $3$, there are published examples with zero data but larger observed
limits~\cite{SUSY-2018-36, atlas_susy_strong_0l}.
The most extreme is $\mathrm{SR}3\ell_{[90,110]}$
of~\cite{SUSY-2018-36}, which reports a background expectation of
$1.08 \pm 0.18$ and expected limits of
$S_{\mathrm{exp}}^{95} = 3.8^{+1.3}_{-0.5}$, but zero observed data with a
whopping $S_{\mathrm{obs}}^{95} = 4.8$.
Using their \pyhf\ models published through \hepdata\ I have failed to
reproduce this effect outside the \histfitter-\roofit ecosystem, but would
not be surprised if similar complications with the test statistic distribution
played a part.

Not only is that limit more than $3$, it is more than the expected limit,
despite the fact that increasing data should increase the observed limit.
The same effect has occurred in our results for DR-OffShell --- its observation
of $21$ data is below the post-fit background is $22.1$, but its observed limit
exceeds the central expected limit.
Another public example of this effect is
$\mathrm{SR}0\textrm{-}\mathrm{ZZ^{tight}_{bveto}}$ of~\cite{SUSY-2018-02}.
\emph{Thanks to Tina~Potter discussions of this problem and for sharing
its details of its modelling.}

Internal discussions of this error led me to perform numerous validation
studies that reproduced this finding with \pyhf\ and with various alternative
constructions of the models.
Various explanations were suggested in these discussions, including problems
with asymmetric systematic variations or asymmetries in the Poisson
distribution.
Fundamentally, a sufficient explanation of this problem is our refusal to
compare models with likelihood ratios.
By instead using \emph{maximum} likelihood ratios, and modifications thereof,
we guarantee that our tests can be suboptimal because they are not protected by
the Neyman–Pearson lemma~\cite{neyman1933ix}.
In this world of model comparison through $p$-values, sub-optimality means
that the test statistic puts some data in the wrong
order~\cite{cousins2008evaluation},
and so by knowingly using suboptimal test statistics we also know that the
results need not behave as if the data had a sensible ordering.
Limits moving in the wrong directions can simply be explained by flawed
test statistics ranking data in a bad order.

@@

It is undoubtedly arguable that the pathological test statistic distributions
observed in DR-Low indicate a bug that should be fixed.
However, the statistical software packages that generated this result
--- \roofit\ through \histfitter ---
define in silicon the standard statistical methods for \atlas\ SUSY searches.
To diverge from them would make our result incomparable.
What they do is the definition of correct.
All \pyhf\ can do is imitate.
There is no such thing as a bug.
